Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2020-09-17 write-up/covid-writeup.tex
Update covid-writeup.tex
switchin to fever inst of symp

The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.  This raises the immediate question whether prevalence estimators can be improved.  In this section, we consider two directions. First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IWP) estimator. We extend the statistical error decomposition to the weighted setting. Second, we consider an extension of compartmental models from infectious disease epidemiology using death data to try and avoid selection bias.  After adjusting for measurement-error, this leads to a model-based forecasts of prevalence. We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators. We provide an appropriate extension of the statistical error decomposition. The method and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.

@@

Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.  Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
\begin{equation}
\label{eq:auxinfoprob}
\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
\end{equation}
Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  This was proposed by~\cite{Chen2019} when covariates were observed on a random sample from the population. Here, we only observe strata information.  Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

@@

Specific weight calculations and adjustments are deferred to Section~\ref{section:applications}. Here, we briefly discuss a non-parametric kernel approach.  We start by using considering a pseudo-likelihood
\begin{equation}
\label{eq:tvpropensity}
\sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} I^{(2)}_{j,t} W_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t} \tilde I_{j,t} \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
\end{equation}

large random samples

covariate information is available on the nonprobability sample at each time $t=1,\ldots,T$;

$$
\sum_{t=1}^T \sum_{i=1}^N K_h(|t^\prime - t|) \tilde W_{j,t} \tilde I_j
$$

We can think of this quantity as a $Z_t$ and each random sample as an estimate $\tilde Z_t$
$$
\| \hat Z_t - Z_t \| = \int \log \left( 1 - \pi_t (x; \theta) \right) dF_t (x)
$$
Where $dF_t$ is the associated measure at time~$t$

We could estimate separately if we had 1 random sample at each.  But perhaps it is better to smooth across; especially

\begin{rmk}[Cumulative demographics]
Most states only report cumulative demographic information.
\end{rmk}

@@

\begin{table}[!th]
\begin{tabular}{c | c | l l | l | l l }
& & \multicolumn{2}{c}{Total Tests} &  & \multicolumn{2}{c}{Positive Tests} \\
\cline{3-4} \cline{6-7}
& & NonRandom & Random & Census & NonRandom & Random \\ \hline
\multirow{2}{*}{Sex$^\star$} & Female & 57\% & 55\% & 49\% & 53\% & - \\
& Male & 43 & 45 & 51 & 47 & -  \\ \hline
\multirow{3}{*}{Age$^\star$} & $<40$ & 47 & 28 & 53 & 47\\
& $40-59$ & 28 & 36 & 25 & 3 0\\
& $\geq 60$ & 25 & 36 & 22 & 13\\ \hline
\multirow{2}{*}{Race$^\star$} & White & 60 & 98 & 85 & 50\\
& Nonwhite & 21 & 8 & 15 & 50 \\ \hline
\multirow{2}{*}{Symptoms$^{\star \star}$} & Yes & 62 & 43 & - & 83 & 62 \\
& No & 38 & 57 & - & 17 & 38  \\ \hline
\multirow{2}{*}{Contact$^{\star \star}$} & Yes & 11 & 1 & - \\
& No & 89 & 99 & - \\ \hline
\end{tabular}
\caption{Temp \\
$\star$: ISDH case-count nonrandom sample \\
$\star\star$: May 2--3 nonrandom sample
}
\end{table}

% Random sample
% Symptom 0, Test 0 2040
% Symptom 0, Test 1 18
% Symptom 1, Test 0 1518
% Symptom 1, Test 1 29
% Total symptom 1; (1518+29)/3605 = 43
% Pos test symptom 1; (29)/(29+18) = 62

% Avg 100 COVID-19 hospitalizations in state in that time
% 650 cases
% percent of cases in nonrandom sample with symptoms = 184/(184+37) = 83%
% Then 83% * 650 = 540 symptomatic cases
% 100/540 = 19% of symptomatic cases were hospitalized

% Example: prior to April 28th, only symptoms and a few others
% 04/11: around 500 cases
% 04/11: 117 hospitalizations -> 615 cases

1. Unknown assumed MAR for Race, Gender, and Age
2. 20\% are asymptomatic (so 78\% + 22\% (20\%) showed no symptoms)

We have a small non-random sample to supplement and get these; then use raking on the larger study.  Appendix XX presents the marginal distributions for the non-random sample per

Heuristically, one may suppose that asymptomatic individuals are less likely to select.  This is born out.

This is not true (need to deal with very bad symptoms cases).

Simplest case we have Age $<40$ in the random sample is $1,017$ and we have $13+17.3+15.2 = 45\% \times 1,002,038$ tests so $450917.1$ tests. Then $N = 6.85$ million while $n = 3658$ leading to $\pi_k = 0.24$.

\begin{rmk}[Use of second nonrandom sample]
The key issue with selection bias is whether the individual is displaying symptoms and had contact with someone who was COVID+.  These are not captured by the

In random sample, $44.2$\% of individuals who tested positive for SARS-CoV-2 positive individuals reported no symptoms during the 2 weeks before testing.

In nonrandom sample, $20.2$\% of individuals who tested positive for SARS-CoV-2 reported no symptoms during the 2 weeks before testing.

>>>>

The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.  This raises the immediate question whether prevalence estimators can be improved.  In this section, we consider two directions. First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IPW) estimator. We extend the statistical error decomposition to the weighted setting. Second, we consider an extension of compartmental models from infectious disease epidemiology using death data to try and avoid selection bias.  After adjusting for measurement-error, this leads to a model-based forecasts of prevalence. We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators. We provide an appropriate extension of the statistical error decomposition. The method and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.

@@

Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.  Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
\begin{equation}
\label{eq:auxinfoprob}
\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
\end{equation}
Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  This was proposed by~\cite{Chen2019} when covariates were observed on a random sample from the population. Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.  We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the design-unbiased estimator
\begin{equation}
\label{eq:auxinfoprob2}
\sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
\end{equation}
Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.  The estimate no relies on random sampling in both the first and second term.

@@

Specific weight calculations and adjustments are deferred to Section~\ref{section:applications}. Here, we briefly discuss a non-parametric kernel approach.  We start by considering a time-varying version of equation~\eqref{eq:auxinfoprob2}
% \begin{equation}
% \label{eq:tvpropensity}
% \sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} I^{(2)}_{j,t} W_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t} \tilde I_{j,t} \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
% \end{equation}
that accounts for random sampling from within the non-random sample.  Note random samples may be quite small at each time while we expect $\pi_t(x;\theta)$ to be a smooth function of time; therefore, we introduce a non-parametric approach to estimation of the selection propensity at time~$t$,
$$
\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} I^{(2)}_{j,t^\prime} W_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
$$
where $K_h$ is a kernel function with tunable parameter $h$. Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time, and s, and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
$$
\hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K(|t^\prime - t|) \frac{n_t}{m} \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T \tilde d_{k,t^\prime}}
$$
where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

@@

\begin{table}[!th]
\begin{tabular}{c | c | r r | r | r r }
& & \multicolumn{2}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
\cline{3-4} \cline{6-7}
& & NonRandom & Random & Census & NonRandom & Random \\ \hline
\multirow{2}{*}{Sex} & Female & 58 & 55 & 51 & 21.7 & 1.4 \\
& Male & 42 & 45 & 49 & 24.2 & 2.1 \\ \hline
\multirow{3}{*}{Age} & $<40$ & 40 & 28 & 53 & 29.7 & 1.7 \\
& $40-59$ & 41 & 36 & 25 & 24.9 & 2.1 \\
& $\geq 60$ & 19 & 36 & 22 & 6.7 & 0.9 \\ \hline
\multirow{2}{*}{Race} & White & 23 & 92 & 87 & 19.5 & 1.5 \\
& Nonwhite & 77 & 8 & 13 & 25.0 & 3.4 \\ \hline
Hispanic & Hispanic & - & - & - & 37.6 & 6.9 \\
Origin & Non-Hispanic & - & - & - & 13.0 & 1.3 \\ \hline
\multirow{2}{*}{Fever} & Yes & 17.0 & 43 & - & 31.6 & 1.9 \\
& No & 38 & 57 & -& 10.4 & 0.9 \\ \hline
\multirow{2}{*}{Household} & Yes & 11 & 1 & - & 46.1 & 24.4 \\
& No & 89 & 99 & - & 21.6 & 1.3 \\ \hline
Prior $+$ & Yes & 6 & 1 & - & 39.2 & 24.4 \\
Test & No & 94 & 99 & - & 21.6 & 1.3 \\ \hline
\end{tabular}
\caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana.}
\end{table}

We correct the nonrandom sample to match the Indiana state-wide prevalence

\begin{subsubsection}{Time-varying doubly-robust prevalence estimation}
